---
title: "ECON621: Introduction and Overview"
author: "Adam Gorski"
date: "Jan. 2, 2020"
output: html_document
---

<style>
div.blue pre { background-color:lightblue; }
div.blue pre.r { background-color:lightgray; }
</style>

Console on the left is for executing code (but not for storing it)

Environment on the top right is the data you have loaded in your current R session

Output windows on the bottom right is where help query responses surface, plots and graphs are displayed

For code you want to keep (anything that isn't an ad hoc fix or experiment), open a script - Cmd/Ctrl + Shift + N

To customize your R Studio, go to Tools > Global Options and change the Appearance (if you'd like)
<div class = "blue">
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Explore some data that is built in to R Studio

```{r , echo = T}
data()
data(mtcars)
```

When `mtcars` appears in the Global Environment, you can click on it to surface an interactive spreadsheet, similar to Excel.

You can explore some of the other datasets to familiarize yourself with R Studio.

Tomorrow we'll start working with and manipulating datasets like this in R.
</div>