---
title: "ECON621: Summarizing and Manipulating Data"
author: "Adam Gorski"
date: "Jan. 8, 2020"
output: html_document
---

<style>
div.blue pre { background-color:lightblue; }
div.blue pre.r { background-color:cornsilk; }
</style>
<div class = "blue">
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today's Lecture:

* Changing Variable Classes

* More Complicated Data Exploration Techniques

* Data Transformation: Casting Wide and Long

Often times, your data analysis will require major modifications to the data itself: changes to the data classes, the granularity, and the shape.

### Changing Variable Classes ###

As we have discussed in previous lectures, every column in a dataframe is a vector, and each vector is a class of data (numeric, text, etc.). If the vector is not the class that you need for your analysis, you can transform the whole vector with an `as.` function. If you have numbers read in as characters, you can convert them using `as.numeric`.
```{r}
df <- data.frame(user = c(letters[1:5]), x = c("3", "1", "5", "3", "8"), stringsAsFactors = F)
class(df$x)
df$x <- as.numeric(df$x)
class(df$x)
```
You can use the same approach to create a new column as well.
```{r}
df <- data.frame(user = c(letters[1:5]), x = c("3", "1", "5", "3", "8"), stringsAsFactors = F)
df$x_as_num <- as.numeric(df$x)
df
```
However, numeric is a less flexible data class than character, so if some of your data doesn't fit that class, R will convert it to `NA`.
```{r}
df <- data.frame(user = c(letters[1:5]), x = c("C", "1", "5", "C", "8"), stringsAsFactors = F)
df$x <- as.numeric(df$x)
df
```
You can similarly modify columns from numeric to character with `as.character`, although this comes up less often.
```{r}
df <- data.frame(user = c(letters[1:5]), x = c(3, 1, 5, 3, 8), stringsAsFactors = F)
class(df$x)
df$x <- as.character(df$x)
class(df$x)
```
Modifying factors is slightly more complicated because R understands the vector as a set of of possible values and assigns each component in the vector an index connected to one of those values. This means that converting the vector to a numeric will return all the values of the indices, *not what the indices represent*, which is what you want. The most straightforward way to do this is to convert factors to character *then convert to numeric*.
```{r}
df <- data.frame(user = c(letters[1:5]), x = c("3", "1", "5", "3", "8"))
class(df$x)
df$x_as_num <- as.numeric(df$x)
df
# We converted a factor straight to a numeric, and look what happened to our values
```
```{r}
df$x_as_num_2 <- as.numeric(as.character(df$x))
df
# Now we have the correct values
```

This actually comes up a lot with R because several defaults in R convert vectors into factors. A good rule to use is to always convert factor variables into characters, unless you want them to stay factors for some reason. You can use `str` to look at the dataframe and `class` to look at each column, but there is a more systematic way to find and convert factors.

The `sapply` function is an iterative function, meaning that it iterates over the data you give it by each vector in the data. What we are interested in is whether each columns is a factor or not. The arguments we need for `sapply` are the data, and the function we are going to apply. Here we want `is.factor`, which just returns a conditional on whether the data class is a factor. (There are several other `is.` functions, such as `is.numeric`.)
```{r}
sapply(df, is.factor)
```
We can use this conditional vector, along with the `sapply` function, to convert factors to characters.
```{r}
df[sapply(df, is.factor)] <- sapply(df[sapply(df, is.factor)], as.character)
sapply(df, class)
```
Once you feel comfortable with approaches like this, you can add them to every script where you are importing data. Generally, it is good practice to standardize your approach to importing, cleaning, and modifying data.

### More Complicated Data Exploration Techniques ###

Understanding an individual variable is useful, but the true power of data science comes from examining the relationships between variables. We'll spend a lot of the course on different approaches to that problem. Now we're going to look at answering some of the data exploration questions from last lecture (min, max, mean, etc.) in the context of other variables in the data.

A quick and useful function for this kind of work is `by`. `by` takes three arguments: the vectors you are analyzing, the vector you are analyzing them by (usually a category variable), and the operation
```{r, include = F}
wine_data <- read.csv("wine_components.csv")
```
```{r}
# If we want to know what the average ash level is for each wine type
by(wine_data$ash, wine_data$wine_type, mean)
```
##### Splitting Data #####
`by` is good for quick answers along one axis. For more complicated data exploration and summarization, especially if the answers themselves will become data for you later, we can use `split` to separate the data, `lapply` to analyze each individual piece, and `ldply` to reassemble the pieces into a useful dataframe.

`split` carves a dataframe up by a variable (usually a category variable) and creates a list of individual dataframes by each of the values in the vector.
```{r}
# Note: here we only limit the columns in the initial dataframe to limit the output
split(wine_data[, 1:5], wine_data$wine_type)
```
The output of the `split` function is a list of dataframes. `lapply` is a function that iterates on every element in a list. So you can wrap `lapply` around `split`.
```{r}
# Here we're looking for the number of rows for each type of wine
lapply(split(wine_data[, 1:5], wine_data$wine_type), nrow)
```
`lapply` gives you the opportunity perform many functions on each dataframe in the list if you write your own function. We'll spend more time practicing building functions later in the course, but this is a good time to start practicing. `function` as a function allows you to choose your own input and output.
```{r}
times2 <- function(x) {x * 2}
times2(14)
```
Within `lapply`, you don't need to name your function, just tell it what to do.
```
# We want a function that counts the rows, finds the average value for malic acid and ash, and returns everything as a dataframe

function(x) {
  data.frame(count = nrow(x), avg_magnesium = mean(x$malic_acid), avg_hue = mean(x$ash))}
```
```{r}
lapply(split(wine_data[, 1:5], wine_data$wine_type), function(x) {
  data.frame(count = nrow(x), avg_magnesium = mean(x$malic_acid), avg_hue = mean(x$ash))})
```
This still leaves you with a list, when a dataframe would be much more convenient. To concatenate (ie, stack) your dataframes, you can use a function called `ldply`, which is part of the `plyr` package.
```
install.packages("plyr")
```
```{r}
library(plyr)
```
`ldply` will combine the elements of a list into a dataframe in whichever way you choose. If your list contains dataframes that have the same column structure, you want to stack them on top of each other. This requires the function `rbind`, which means 'row bind.'

You can wrap your `ldply` function around your `lapply` to do the whole process in one step.
```{r}
ldply(lapply(split(wine_data[, 1:5], wine_data$wine_type), function(x) {
  data.frame(count = nrow(x), avg_magnesium = mean(x$malic_acid), avg_hue = mean(x$ash))}), rbind)

# We can review this process step by step:

# 1. How do we want to split our data? In this case, by wine type
wine_type_list <- split(wine_data[, 1:5], wine_data$wine_type)

wine_type_list
# 2. What do we want to do with each separate data frame?
wine_type_list <- lapply(wine_type_list, function(x) {
  data.frame(count = nrow(x), avg_magnesium = mean(x$malic_acid), avg_hue = mean(x$ash))})

wine_type_list
# 3. How do we want to combine our results? In this case, by row
ldply(wine_type_list, rbind)
```
This nested functions approach to reorienting data is very powerful and flexible. You can split data by multiple columns. The function you build into your `lapply` function can perform dozens of operations using the data you feed it. This approach is also fast and efficient: instead of creating and storing intermediate tables, you are left only with the desired output.

### Data Transformation: Casting Wide and Long ###
There are two very distinct data shapes: wide and long. Wide is columns at the expense of rows; long is rows at the expense of columns.
```{r}
# A wide data example
data.frame(school = c("Roosevelt Elementary", "Jefferson Elementary"),
           grade_1_all = c(67, 74), grade_1_f = c(68, 77), grade_1_m = c(66, 75),
           grade_2_all = c(62, 73), grade_2_f = c(66, 76), grade_2_m = c(64, 74))
# A long data example
data.frame(school = c(rep("Roosevelt Elementary", 6), rep("Jefferson Elementary", 6)),
                      grade = c(1, 1, 1, 2, 2, 2), gender = c("all", "f", "m", "all", "f", "m"),
                      proficiency = c(67, 68, 66, 62, 66, 64, 74, 77, 75, 73, 76, 74))
```
There are advantages to both shapes. Wide data is easier to understand visually, easier to calculate statistics along columns and rows, and better for regression analysis. Long data requires less storage capacity (fewer empty values) and is faster and more efficient for most data management.

Casting data will take practice to understand conceptually, but it is a crucial component to understanding which data structures you need and how to build them.

Converting (ie, casting) wide data long or long data wide is frequently necessary for modifying external datasets and for working with your own data to run regressions or create visualizations. A good package for these operations is `reshape2`.
```
install.packages("reshape2")
```
```{r}
library(reshape2)
```
##### Long to Wide #####
Casting data long to wide is common in data science because you are often retrieving data that is stored in long format but that you need to convert to wide format for analysis. `dcast` is a good function for doing so. After providing your dataframe, you construct a formula that is: `id ~ variable`. `id` is the row name, `variable` is the column name. The argument `value.var` tells the function what values to use to populate the dataframe.
```{r}
school_long <- data.frame(school = c("Roosevelt Elementary","Roosevelt Elementary",
                                     "Jefferson Elementary", "Jefferson Elementary"),
                          level = c("grade_1", "grade_2", "grade_1", "grade_2"),
                          score = c(67, 62, 74, 73))
school_long
dcast(school_long, school ~ level, value.var = "score")
```
##### Wide to Long #####
`melt` is the function (also the verb) for casting long data wide. Think of a melting piece of cheese pizza, growing longer as the cheese stretches downward. `melt` is oriented slightly differently than `dcast`, but the same components are used. The `melt` function considers a dataset to consist of two types of variables: id variables and measure variables. You tell the function which variable(s) is an id, and it will consider the rest to be measure values described by their column name. It will create a row for each unique combination of id and measure variable. In the `melt` function, you can use the `variable.name` argument to name the column where the previous column names will be stored, and you can use the `value.name` argument to name the column where the measure values will be stored.
```{r}
school_wide <- data.frame(school = c("Roosevelt Elementary", "Jefferson Elementary"),
                          grade_1 = c(67, 74), grade_2 = c(62, 73))
school_wide
melt(school_wide, id = "school") # Variable and value names not provided
melt(school_wide, id = "school", variable.name = "level", value.name = "score")
```
</div>